---
title: "Assignment 5 - Bay Area Bike Rental Operation Data Analysis Report"
author: "Imran Rhemtulla"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis of the Weather Dataset

Pre-processing steps were applied before data analyses began. “T” values representing trace
amounts of precipitation within the precipitation_inches variable were converted to 0 and the variable was converted to numerical. “rain” was converted to “Rain” in the events variable to
properly group the variable. Within the events variable, there were 1473 observations that were
null. These observations were converted to “No Event”, indicating that no weather event took
place. The variable zip_code was converted to a categorical variable from numerical as there were
only 5 zip codes, one for each of the cities described in the dataset.

Within the weather dataset there are a total of 1825 observations and 15 variables. 5
(0.27%) observations for each date were present in the data set and ranged from 1/1/2014 to
6/15/2014. In the city variable there were 365 (20%) observations from Mountain View, 365 (20%)
observations from Palo Alto, 365 (20%) observations from Redwood City, 365 (20%) observations
from San Francisco, and 365 (20%) observations from San Jose (Figure 1). Similarly, the zip codes
within the zip_codes variable were general and represent each of the regions. 365 (20%)
observations were present for 94041 (Mountain View), 365 (20%) observations for 94301 (Palo
Alto), 365 (20%) observations for 94063 (Redwood City), 365 (20%) observations for 94107 (San
Francisco), and 365 (20%) observations for 95113 (San Jose) (Figure 1) Of these dates, 1473
(80.71%) were recorded as undergoing no weather event across all cities, 282 (15.45%) were noted
to experience rain, 57 (3.12%) had fog documented across all cities, and 13 (0.71%) had both fog
and rain.

Within the numerical variables, 1616 (88.55%) of observations for precipitation_inches
were zero. 451 observations for maximum gust speed were missing (24.71%), thus maximum gust
speed had less than 80% of all values as non-NA values. The mean of maximum temperature was
271.03 with a standard deviation (SD) of 8.26. The mean of mean temperature was 62.04 with an SD of 6.75. The mean of minimum temperature was 52.83 with an SD of 6.67. The mean of max visibility was 10.86 with an SD of 2.62. The mean of mean visibility was 9.97 with an SD of 1.62. The mean of minimum visibility was 8.11 with an SD of 3.04. The mean of maximum wind speed 16.43 was with an SD of 7.32. The mean of mean wind speed was 6.11 with an SD of 3.05. The mean of maximum wind speed was 22.69 with an SD of 9.09. Precipitation had an extremely high variation of coefficient of 5.97 and broad 98-percentile range of 9-45.81, indicating the presence of outliers. The mean precipitation was 0.03 with an SD of 0.18 as well. The variable describing cloud cover had a mean of 3, with an SD of 2.30. Representative histograms can be found in Figure 2.

```{r cars}
library(funModeling)
library(tidyverse)
library(Hmisc)
library(stringr)
library(dplyr)

# Import weather dataset as a dataframe
weather_data <- read.csv("weather.csv", header = TRUE, sep = ",")

# Head the first observations and view the total number of observations and variables
glimpse(weather_data)
  # 1825 total observations
  # 15 variables
  # Data is in imperial measurements (F, miles, mph, inches)

# Replace T values in precipitation_inches with 0 and convert to numeric
weather_data$precipitation_inches <- stringr::str_replace(weather_data$precipitation_inches, "T", "0")
weather_data <- weather_data %>% mutate(precipitation_inches = as.numeric(precipitation_inches))

# Replace empty values with "No Event" in events
weather_data$events[weather_data$events == ""] <- "No Event"
weather_data$events[weather_data$events == "rain" | weather_data$events == "Rain-Thunderstorm"] <- "Rain"

# Convert zip_code to character 
weather_data$zip_code <- as.character(weather_data$zip_code)

# Gain metrics on data types, zeroes, infinite numbers, and missing values
print(status(weather_data))
  # 451 missing values in max_gust_speed_mph
  # 1543 0s in precipitation_inches (no precipitation on that day), will skew histograms heavily
  # Variables to turn into factors: cloud_cover, events, city
  # Variables of interest: precipitation_inches (contains T and integer values)
  # zip_code has as many unique values as city (zipcode is not unique)

# Analyze categorical variables
freq(weather_data) 
  # Equal distribution entires for all 5 cities
  # 1473 NULL values in events variable
  # precipitation_inches is treated as a categorical variable due to T 

  #Export frequency plots to current directory
  freq(weather_data, path_out = ".")
     
print(profiling_num(weather_data))
  # Right skew: max_wind_speed_mph, max_gust_speed_mph (also have high st dev)
  # wind speed descriptor variables have high variation_ceof

plot_num(weather_data)
  # max_temp_f -> unimodal with potentially some outliers
  # mean_temp_f -> unimodal
  # min_temp_f -> unimodal
  # max_visibility_miles -> one large peak, will have high std dev
  # mean_visibility miles -> one large peak, but balanced on either side
  # min_visibility_miles -> left skew, outliers present 15+
  # max_wind_speed_mph -> outliers present, towards 120
  # mean_wind_speed -> slight right skew, potential outliers towards 20
  # max_gust_speed_mph -> right skew, definitely has outliers towards 90
  # cloud_cover -> should be treated as factor, but well distributed
  # zip_code -> being treated as numerical, should convert to a factor
  #Export numerical plots to current directory
  plot_num(weather_data, path_out = ".")

describe(weather_data)
  # Consistent observations with above

saveRDS(weather_data, "weather_processed.rds")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
